Constant Partying: \\ Introducing a Generic Toolkit for Recursive Partitioning in R

### 2014-02-10 pkg/partykit/inst/NUS-2014/constparty_TH_AZ.Rnw
show example tree algorithm
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@456 255f4249-2263-4417-b1ea-8498779e27a7

<<cache>>=
if(file.exists("slides-trees.rda")) load("slides-trees.rda") else {
<<rpart>>
<<ctree>>
<<evtree>>
<<pmml>>
<<mob>>
save(rp, ct, ev, pm, mb, file = "slides-trees.rda")
}
@

@@

\begin{frame}
\frametitle{Your own tree}

>>>>

<<mytree, eval=FALSE>>=
findsplit <- function(response, data, weights) {

  ### extract response values from data
  y <- data[[response]]

  ### cycle through all features
  logpmin <- 0; xselect <- NULL
  for (i in which(names(data) != response)) {

    ### expand data by case weights
    x <- data[[i]]
    xt <- rep(x, weights) 
    yt <- rep(y, weights)

    ### set-up all possible splits in two daughter nodes
    lev <- levels(xt[drop = TRUE])
    if (length(lev) == 1) next()
    if (length(lev) == 2) comb <- list(lev[1])
    if (length(lev) > 2) 
        comb <- do.call("c", lapply(1:(length(lev) - 2),
            function(x) combn(lev,x, simplify = FALSE)))

    ### assess all potential splits by their chi^2 test log p-value
    logp <- sapply(comb, function(q) {   
      ct <- chisq.test(table(yt, xt %in% q))
      pchisq(ct$statistic, ct$parameter, log = TRUE, lower.tail = FALSE)
    })

    ### if the best split in variable i significant AND
    ### better than what we already had store this information
    if (min(logp) < logpmin & min(logp) < log(0.01)) {
      logpmin <- min(logp)
      xselect <- i
      splitpoint <- comb[[which.min(logp)]]
    }
  }

  ### no significant split found, give up  
  if (is.null(xselect)) return(NULL)

  ### return split as partysplit object
  return(partysplit(
      varid = as.integer(xselect),         ### which variable?   
      index = levels(data[[xselect]]) %in%
              splitpoint + 1L,             ### which split point?
      info = list(p.value = exp(logpmin)   ### save p-value in addition
  )))
}
growtree <- function(id = 1L, response, data, weights) {

  ### for less than 30 observations stop here
  if (sum(weights) < 30) return(partynode(id = id))

  ### find best split
  sp <- findsplit(response, data, weights)
  ### no split found, stop here
  if (is.null(sp)) return(partynode(id = id))

  ### actually split the data
  kidids <- kidids_split(sp, data = data)

  ### set-up all daugther nodes
  kids <- vector(mode = "list", length = max(kidids))
  for (kidid in 1:max(kidids)) {
  ### select observations for current node
  w <- weights
  w[kidids != kidid] <- 0
  ### get next node id
  if (kidid > 1) {
    myid <- max(nodeids(kids[[kidid - 1]]))
  } else {
    myid <- id
  }
  ### start recursion on this daugther node
  kids[[kidid]] <- growtree(id = as.integer(myid + 1), response, data, w)
  }

  ### return nodes  
  return(partynode(id = as.integer(id), split = sp, kids = kids))
}
mytree <- function(formula, data, weights = NULL) {

  ### name of the response variable
  response <- all.vars(formula)[1]
  ### data without missing values, response comes last
  data <- data[complete.cases(data), c(all.vars(formula)[-1], response)]
  ### data is factors only
  stopifnot(all(sapply(data, is.factor)))

  if (is.null(weights)) weights <- rep(1L, nrow(data))
  ### weights are case weights, i.e., integers
  stopifnot(length(weights) == nrow(data) &
    max(abs(weights - floor(weights))) < .Machine$double.eps)

  ### grow tree
  nodes <- growtree(id = 1L, response, data, weights)

  ### compute terminal node number for each observation
  fitted <- fitted_node(nodes, data = data)
  ### return rich constparty object
  ret <- party(nodes, 
    data = data,
    fitted = data.frame(
      "(fitted)" = fitted,
      "(response)" = data[[response]],
      "(weights)" = weights,
      check.names = FALSE),
    terms = terms(formula))
  as.constparty(ret)
}  
myttnc <- mytree(Survived ~ Class + Age + Gender, data = ttnc)
@

<<cache>>=
if(file.exists("slides-trees.rda")) load("slides-trees.rda") else {
<<rpart>>
<<ctree>>
<<evtree>>
<<pmml>>
<<mob>>
<<mytree>>
save(rp, ct, ev, pm, mb, myttnc, file = "slides-trees.rda")
}
@

@@

\begin{frame}
\frametitle{Your first partykit tree}

Finally, we want to set-up a little program that implements the
following tree algorithm for the Titanic data
\begin{itemize}
\item In each node, do
  \begin{itemize}
    \item For each explanatory variable, do
    \begin{itemize}
      \item compute the minimal $\chi^2$ $p$-value over all
            possible splits in two groups
    \end{itemize}
  \item Select the best variable/split combination
        and implement the split
  \end{itemize}
\item Recurse until $p$-value > .01 or sample size too small.
\end{itemize}

We only need three little functions:
\begin{itemize}
\item \code{findsplit}
\item \code{growtree}
\item \code{mytree}
\end{itemize}

\begin{frame}
\frametitle{Your first partykit tree}

\setkeys{Gin}{width=1.1\textwidth}
\hspace*{-0.4cm}%
<<Titanic-mytree-vis, fig = TRUE, height=8, width=12>>=
plot(myttnc)
@
